Stochastic Variational Multioutput Gaussian Processes

### 2014-03-07 paper/experiments.tex
updated paper
clean figures

\subsection{TOY PROBLEM}
In this toy problem we use two related outputs generated from the same latent function $sin(x)$ but are corrupted by independent noises: $y_1(x) = sin(x) + \epsilon$ and $y_2(x) = -sin(x) + \epsilon$, $\epsilon \sim \Normal(0,0.01)$.
Each output has missing values in one region of the input space.
% Say that this is from a single gpsvi
The predictive distributions of the model compared to that of independent GPs with stochastic variational inference are shown in figure \ref{fig:toy}.
Furthermore, the inference algorithm correctly learns the correlation between the two functions giving weights $w_{11} = 1.07$ and $w_{21} = -1.06$.

\begin{figure*}
\centering
\begin{tabular}{cccc}
\includegraphics[scale=0.25]{figures/toy-slfm-y1.eps} &
\includegraphics[scale=0.25]{figures/toy-svigp-y1.eps} &
\includegraphics[scale=0.25]{figures/toy-slfm-y2.eps} &
\includegraphics[scale=0.25]{figures/toy-svigp-y2.eps}
\end{tabular}
\caption{Predictive distributions of the multioutput GPs (first and third figure) and independent GPs using stochastic variational inference (second and last figure) for the  toy problem. Solid black line: predictive mean; grey bar: two standard deviations; magenta dots: real observations; blue dots: missing data. The black crosses show the locations of the inducing inputs.}
\label{fig:toy}
\end{figure*}

\subsection{EVALUATION OF THE PREDICTIVE ACCURACY}
\subsubsection{Foregin Exchange Rate Prediction}
In this experiment we predict the foreign exchange rate.
The smse are shown in Table \ref{tab:fx}.

\begin{table}[h]
\caption{Performance comparison by SMSE on the fx dataset.}
\label{tab:fx}
\begin{center}
\begin{tabular}{c|c}
method & smse \\ \hline
SLFM (Q=2) & 0.2232 \\
SLFM (Q=1) & 0.48 \\
CGP & 0.2795 \\
ICM & 0.3927
\end{tabular}
\end{center}
\end{table}

@@

\subsubsection{Weather Data Prediction}

\begin{figure*}
\centering
\begin{tabular}{cc}
\includegraphics[scale=0.3]{figures/weatherCambermet.eps} &
\includegraphics[scale=0.3]{figures/weatherCambermet.eps} \\
\includegraphics[scale=0.3]{figures/weatherChimet.eps} &
\includegraphics[scale=0.3]{figures/weatherChimet.eps}
\end{tabular}
\caption{Predictive distributions of the multioutput GPs (left column) and full independent GPs (right column) for the air temperature problem.}
\label{fig:toy}
\end{figure*}

\subsection{LARGE-SCALE EXPERIMENTS}
\subsubsection{Robort Arms}
\subsubsection{Foreign Exchange Rate Prediction}

>>>>

\subsection{TOY PROBLEM}
In this toy problem we simulate two related outputs from the same latent function $sin(x)$ corrupted by independent noises: $y_1(x) = sin(x) + \epsilon$ and $y_2(x) = -sin(x) + \epsilon$, $\epsilon \sim \Normal(0,0.01)$.
Each output has missing values in one region of the input space.
% Say that this is from a single gpsvi
The predictive distributions of the model compared to that of independent GPs with stochastic variational inference are shown in figure \ref{fig:toy}.
Furthermore, the inference algorithm correctly learns the correlation between the two functions giving weights $w_{11} = 1.07$ and $w_{21} = -1.06$.

\begin{figure*}
\centering
\begin{tabular}{cccc}
\includegraphics[scale=0.2]{figures/toy-slfm-y1.eps} &
\includegraphics[scale=0.2]{figures/toy-svigp-y1.eps} &
\includegraphics[scale=0.2]{figures/toy-slfm-y2.eps} &
\includegraphics[scale=0.2]{figures/toy-svigp-y2.eps}
\label{fig:toy}
\end{tabular}
\caption{Predictive distributions of the multioutput GPs (first and third figure) and independent GPs using stochastic variational inference (second and last figure) for the  toy problem. Solid black line: predictive mean; grey bar: two standard deviations; magenta dots: real observations; blue dots: missing data. The black crosses show the locations of the inducing inputs.}
\label{fig:toy}
\end{figure*}

\subsection{FOREIGN EXCHANGE RATE PREDICTION}
In this experiment we predict the foreign exchange rate.
The smse are shown in Table \ref{tab:fx}.

\begin{table}[h]
\caption{Performance comparison by SMSE on the fx dataset.}
\label{tab:fx}
\begin{center}
\begin{tabular}{c|c}
method & smse \\ \hline
SLFM (Q=2) & 0.2232 \\
SLFM (Q=1) & 0.48xx \\
CGP & 0.2795 \\
ICM & 0.3927
\end{tabular}
\end{center}
\end{table}

@@

\subsection{AIR TEMPERATURES PREDICTION}

\begin{table}[h]
\caption{Performance comparison by SMSE on the air temperature dataset.}
\label{tab:fx}
\begin{center}
\begin{tabular}{c|c}
method & smse \\ \hline
SLFM & 0.0680 \\
CGP & 0.0994 \\
IGP & 0.8944
\end{tabular}
\end{center}
\end{table}

\begin{figure*}
\centering
\begin{tabular}{cc}
\includegraphics[scale=0.3]{figures/slfm-weatherCambermet.eps} &
\includegraphics[scale=0.3]{figures/weatherCambermet.eps} \\
\includegraphics[scale=0.3]{figures/slfm-weatherChimet.eps} &
\includegraphics[scale=0.3]{figures/weatherChimet.eps}
\end{tabular}
\caption{Predictive distributions of the multioutput GPs (left column) and full independent GPs (right column) for the air temperature problem.}
\label{fig:weather}
\end{figure*}

\subsection{ROBOT ARMS}

### 2014-03-07 paper/inference.tex
clean figures

\subsection{Variational Inference in Sparse GPs Revisited}
In this section we review inference in sparse GPs (as presented in Titsias).
The posterior in an augmented model is $p(\f, \u | \y) = p(\f | \u, \y) p(\u | \y)$.
The key property of such augmented GPs  is the notation of \textit{sufficient statistics}: given the inducing points $\u$, the latent values $\f$ are independent with any other set of latent values (e.g. the test set).
In the optimal setting when $\u$ is the sufficient statistics of $\f$, it should hold that $p(\f | \u, \y) = p(\f | \u)$ as $\y$ is only the noisy version of $\f$.
This leads to choosing a variational approximation of the posterior which factorizes as $q(\f, \u | \y) = p(\f | \u) q(\u | \y)$.
Since the conditional $p(\f | \u)$ is known, variational inference becomes learning an optimal posterior $q(\u | \y)$ only. \\

@@

\noindent
In variational inference, this is done by optimizing the evidence lower bound (ELBO) of the log marginal:
\begin{align}
%\nonumber
%\log p(\y) \ge& \int q(\u, \v) \log \frac{p(\y | \u, \v) p(\u, \v)}{q(\u, \v)} \der \u \der \v \\
%\nonumber
%=& \int q(\u, \v) \log p(\y | \u, \v)  \der \u \der \v 
%+ \int q(\u, \v) \log \frac{p(\u, \v)}{q(\u, \v)} \der \u \der \v \\
\nonumber
&\log p(\y) \ge \int q(\u, \v) \log p(\y | \u, \v)  \der \u \der \v \\
&- \sum_{j=1}^Q \KL[q(\u_j) || p(\u_j)] - \sum_{i=1}^P \KL[q(\u_i) || p(\u_i)],
\label{eq:elbo}
\end{align}
which is derived using Jensen's inequality and the fact that both of $q(\u, \v)$ and $p(\u, \v)$ fully factorize.
Since $q(\u_j), q(\v_i), p(\u_j), p(\v_i)$ are all multivariate Gaussian distributions, the KL divergences are analytically tractable and require $\calO(M^3)$ computation. To compute the expected likelihood term in the ELBO we first see that
\begin{align}
\nonumber
\log \text{ } &p(\y | \u, \v)
% &= \log \Eb{p(\y | \g, \h)}_{p(\g,\h | \u, \v)} \\
% \nonumber
\ge \Eb{\log p(\y | \g, \h)}_{p(\g,\h | \u, \v)}  \\
&= \sum_{i=1}^P \sum_{n=1}^N \Eb{\log p(y_{in} | \g_n, h_{in}) }_{p(\g_n | \u) p(\h_{in} | \v_i)} 
\end{align}
where $\g_n = \{g_{jn} = (\g_j)_n\}_{j=1}^Q$.
The inequality is due to Jensen's inequality and the equality is due to the fact that the likelihood fully factorizes.
Each individual term $l_{in} \define \Eb{\log p(y_{in} | \g_n, h_{in}) }_{p(\g_n | \u) p(\h_in | \v_i)}$ is computed using the identity in eq. \ref{eq:identity} and  given in eq \ref{eq:lin} (both are in the Appendix).
\newcommand{\Ahi}{\A^h_i}
\newcommand{\Zi}{\Z_i}
 Substituting $l_{in}$ into equation \ref{eq:elbo} and carrying out the integral using the identity in eq. \ref{eq:identity} again we get:
\begin{align}
\nonumber
&\log p(\y)
\ge \sum_{i,n}
\bigg( \log  \Normal(y_{in}; \tilde{\mu}_{in}, \beta_i^{-1})
          - \frac{1}{2} \beta_i \sum_{j=1}^Q w_{ij}^2 \tilde{k}_{nn} \\ \nonumber
         &- \frac{1}{2} \beta_i \tilde{k}^h_{inn}
         - \frac{1}{2} \beta_i \trace  \sum_{j=1}^Q w_{ij}^2 \S_j \mat{\Lambda}_{jn} - \beta_i \frac{1}{2} \trace \S^h_i \mat{\Lambda}_{in} 
\bigg) \\
&- \sum_{j=1}^Q \KL[q(\u_j) || p(\u_j)] - \sum_{i=1}^P \KL[q(\v_i) || p(\v_i)]  \define \calL,
\end{align}
where, with the help of the auxiliary matrices $\A_j = k(\X,\Zj)k(\Zj,\Zj)^{-1}$ and  
$\Ahi = k(\X_i,\Zhi)k(\Zhi,\Zhi)^{-1}$, 
\begin{align}
\tilde{\mu}_{in}
%&= \sum_{j=1}^Q w_{ij} k(\x_n, \Zj)k(\Zj,\Zj)^{-1}\m_j + k(\x_n, \Zhi)k(\Zhi,\Zhi)^{-1}\mhi \\
&= \sum_{j=1}^Q w_{ij} \A_j(n,:) \m_j + \Ahi(n,:) \mhi \\
\mat{\Lambda}_{jn}
%&= k(\Zj,\Zj)^{-1} k(\Zj, \x_n) k(\x_n, \Zj) k(\Zj,\Zj)^{-1} \\
&= \A_j(n,:)^T \A_j(n,:)  \\
\mat{\Lambda}_{in}
%&= k(\Zhi,\Zhi)^{-1} k(\Zhi, \x_n) k(\x_n, \Zhi) k(\Zhi,\Zhi)^{-1}
&= \Ahi(n,:)^T \Ahi(n,:) 
\end{align}
where $\tilde{k}_{nn} = (\tilde{\K})_{nn}$, $\tilde{k}^h_{inn} = (\tilde{\K}^h_i)_{nn}$, $\mu_{jn} = (\Mu_j)_n$,  $\mu^h_{in} = (\Mu^h_i)_n$, and $\A_j(n,:)$ selects the $n$-th row vector of $\A_j$.

% More on this elbo term
\noindent Notice that this ELBO clearly generalizes the standard GP regression. In particular, setting $P = Q = 1$, $w_i = 1$ and $h_i(\x) = 0$ we recover the bound in Hensman et al \cite{hensmangaussian}.
Due to the decomposition of this bound, we can use stochastic gradient descent to learn the variational parameters.

>>>>

\subsection{Variational Inference in Sparse GPs Revisited}
In this section we review inference in sparse GPs (as presented in Titsias).
The posterior in an augmented model is $p(\f, \u | \y) = p(\f | \u, \y) p(\u | \y)$.
The key property of such augmented GPs  is the notion of \textit{sufficient statistics}: given the inducing points $\u$, the latent values $\f$ are independent with any other set of latent values (e.g. the test set).
In the optimal setting when $\u$ is the sufficient statistics of $\f$, it should hold that $p(\f | \u, \y) = p(\f | \u)$ as $\y$ is only the noisy version of $\f$.
This leads to choosing a variational approximation of the posterior which factorizes as $q(\f, \u | \y) = p(\f | \u) q(\u | \y)$.
Since the conditional $p(\f | \u)$ is known, variational inference becomes learning an optimal posterior $q(\u | \y)$ only. \\

@@

\noindent
In variational inference, this is done by optimizing the evidence lower bound (ELBO) of the log marginal:
\begin{align}
%\nonumber
%\log p(\y) \ge& \int q(\u, \v) \log \frac{p(\y | \u, \v) p(\u, \v)}{q(\u, \v)} \der \u \der \v \\
%\nonumber
%=& \int q(\u, \v) \log p(\y | \u, \v)  \der \u \der \v 
%+ \int q(\u, \v) \log \frac{p(\u, \v)}{q(\u, \v)} \der \u \der \v \\
\nonumber
&\log p(\y) \ge \int q(\u, \v) \log p(\y | \u, \v)  \der \u \der \v \\
&- \sum_{j=1}^Q \KL[q(\u_j) || p(\u_j)] - \sum_{i=1}^P \KL[q(\v_i) || p(\v_i)],
\label{eq:elbo}
\end{align}
which is derived using Jensen's inequality and the fact that both of $q(\u, \v)$ and $p(\u, \v)$ fully factorize.
Since $q(\u_j), q(\v_i), p(\u_j), p(\v_i)$ are all multivariate Gaussian distributions, the KL divergences are analytically tractable and require $\calO(M^3)$ computation. To compute the expected likelihood term in the ELBO we first see that
\begin{align}
\nonumber
\log \text{ } &p(\y | \u, \v)
% &= \log \Eb{p(\y | \g, \h)}_{p(\g,\h | \u, \v)} \\
% \nonumber
\ge \Eb{\log p(\y | \g, \h)}_{p(\g,\h | \u, \v)}  \\
&= \sum_{i=1}^P \sum_{n=1}^N \Eb{\log p(y_{in} | \g_n, h_{in}) }_{p(\g_n | \u) p(\h_{in} | \v_i)} 
\end{align}
where $\g_n = \{g_{jn} = (\g_j)_n\}_{j=1}^Q$.
The inequality is due to Jensen's inequality and the equality is due to the fact that the likelihood fully factorizes.
Each individual term $l_{in} \define \Eb{\log p(y_{in} | \g_n, h_{in}) }_{p(\g_n | \u) p(\h_in | \v_i)}$ is computed using the identity in eq. \ref{eq:identity} and  given in eq \ref{eq:lin} (both are in the Appendix).
\newcommand{\Ahi}{\A^h_i}
\newcommand{\Zi}{\Z_i}
 Substituting $l_{in}$ into equation \ref{eq:elbo} and carrying out the integral using the identity in eq. \ref{eq:identity} again we get:
\begin{align}
\nonumber
&\log p(\y)
\ge \sum_{i,n}
\bigg( \log  \Normal(y_{in}; \tilde{\mu}_{in}, \beta_i^{-1})
          - \frac{1}{2} \beta_i \sum_{j=1}^Q w_{ij}^2 \tilde{k}_{nn} \\ \nonumber
         &- \frac{1}{2} \beta_i \tilde{k}^h_{inn}
         - \frac{1}{2} \beta_i \trace  \sum_{j=1}^Q w_{ij}^2 \S_j \mat{\Lambda}_{jn} - \beta_i \frac{1}{2} \trace \S^h_i \mat{\Lambda}_{in} 
\bigg) \\
&- \sum_{j=1}^Q \KL[q(\u_j) || p(\u_j)] - \sum_{i=1}^P \KL[q(\v_i) || p(\v_i)]  \define \calL,
\end{align}
where, with the help of auxiliary matrices $\A_j = k(\X,\Zj)k(\Zj,\Zj)^{-1}$ and  
$\Ahi = k(\X_i,\Zhi)k(\Zhi,\Zhi)^{-1}$, 
\begin{align}
\tilde{\mu}_{in}
%&= \sum_{j=1}^Q w_{ij} k(\x_n, \Zj)k(\Zj,\Zj)^{-1}\m_j + k(\x_n, \Zhi)k(\Zhi,\Zhi)^{-1}\mhi \\
&= \sum_{j=1}^Q w_{ij} \A_j(n,:) \m_j + \Ahi(n,:) \mhi \\
\mat{\Lambda}_{jn}
%&= k(\Zj,\Zj)^{-1} k(\Zj, \x_n) k(\x_n, \Zj) k(\Zj,\Zj)^{-1} \\
&= \A_j(n,:)^T \A_j(n,:)  \\
\mat{\Lambda}_{in}
%&= k(\Zhi,\Zhi)^{-1} k(\Zhi, \x_n) k(\x_n, \Zhi) k(\Zhi,\Zhi)^{-1}
&= \Ahi(n,:)^T \Ahi(n,:) 
\end{align}
where $\tilde{k}_{nn} = (\tilde{\K})_{nn}$, $\tilde{k}^h_{inn} = (\tilde{\K}^h_i)_{nn}$, $\mu_{jn} = (\Mu_j)_n$,  $\mu^h_{in} = (\Mu^h_i)_n$, and $\A_j(n,:)$ selects the $n$-th row vector of $\A_j$.

% More on this elbo term
\noindent Notice that this ELBO clearly generalizes the standard GP regression. In particular, setting $P = Q = 1$, $w_i = 1$ and $h_i(\x) = 0$ we recover the bound in Hensman et al \citet{hensmangaussian}.
Due to the decomposition of this bound, we can use stochastic gradient descent to learn the variational parameters.

### 2014-03-07 paper/model.tex
clean figures

\noindent Consider the semiparametric latent factor models (SLFM) \cite{teh-et-al-aistats-05}.
It assumes $Q$ basic independent processes, $g_j(\x) \sim \GP(0, k(\x,\x';\vectheta_j))$, where $\vectheta_j$ is the hyperparameters of the covariance function of $g_j(\x)$.
Each (latent) output is a weighted combination of these basic processes, i.e.
\begin{align}
f_i(\x) = \sum_{j=1}^Q w_{ij} g_j(\x), \quad i = 1 \hdots P,
\end{align}
In \cite{seeger2005semiparametric}, an extra degree of freedom is given by allowing each output $i$ to have its own process $h_i(\x) \sim \GP(0, k(\x,\x'; \vectheta^h_i)$ leading to
\begin{align}
f_i(\x) = \sum_{j=1}^Q w_{ij} g_j(\x) + h_i(\x), 
\end{align}
We emphasize that, although $k$ denotes covariance functions in general, each independent process can have its own covariance function.

@@

\noindent
\textbf{Likelihood}
The likelihood as usual follows the standard iid Gaussian likelihood,
\begin{align}
p(\y | \g, \h ) = \prod_{i=1}^P p( \y_i ; \g, \h_i) = \prod_{i=1}^P \Normal( \y_i ; \sum_{j=1}^Q w_{ij} \g_j + \h_i, \beta_i^{-1} \I).
\end{align}

>>>>

\noindent Consider the semiparametric latent factor models (SLFM) \citet{teh-et-al-aistats-05}.
It assumes $Q$ basic independent processes, $g_j(\x) \sim \GP(0, k(\x,\x';\vectheta_j))$, where $\vectheta_j$ is the hyperparameters of the covariance function of $g_j(\x)$.
Each (latent) output is a weighted combination of these basic processes, i.e.
\begin{align}
f_i(\x) = \sum_{j=1}^Q w_{ij} g_j(\x), \quad i = 1 \hdots P,
\end{align}
In \citet{seeger2005semiparametric}, an extra degree of freedom is given by allowing each output $i$ to have its own process $h_i(\x) \sim \GP(0, k(\x,\x'; \vectheta^h_i)$ leading to
\begin{align}
f_i(\x) = \sum_{j=1}^Q w_{ij} g_j(\x) + h_i(\x), 
\end{align}
We emphasize that, although $k$ denotes covariance functions in general, each independent process can have its own covariance function.

@@

\noindent
\textbf{Likelihood}
The likelihood as usual follows the standard iid Gaussian likelihood,
\begin{align}
p(\y | \g, \h ) = \prod_{i=1}^P \Normal( \y_i ; \sum_{j=1}^Q w_{ij} \g_j + \h_i, \beta_i^{-1} \I),
\end{align}
where each term in the product is the likelihood of an output, $p( \y_i ; \g, \h_i)$.
